\documentclass[a4paper, oneside, 10pt, draft]{memoir}
\input{packages}
\chapterstyle{culver}
\usepackage{fixme}
\usepackage[english]{babel}
%\usepackage[utopia]{mathdesign}

\usepackage[osf,sc]{mathpazo}
\linespread{1.05}
%\usepackage{fourier}


\usepackage{semantic}
\author{Jesper Louis
  Andersen\\jesper.louis.andersen@gmail.com\\140280-2029}
\title{Lambda}
\date{\today}

\newlength{\drop}
\newcommand*{\titleM}{\begingroup% Misericords, T&H p 153
  \drop = 0.08\textheight
  \centering
  {\Huge\bfseries Lambda}\\[\baselineskip]
  {\scshape IR of exsml}\\[\baselineskip]
  {\scshape by}\\[\baselineskip]
  {\large\scshape Jesper Louis Andersen\\jesper.louis.andersen@gmail.com}\par
  \endgroup}

\bibliographystyle{plain}

\begin{document}
\newcommand{\janusz}{$\mathrm{JANUS}_0$}
\newcommand{\januso}{$\mathrm{JANUS}_1$}
\newcommand{\lift}[1]{\lfloor #1 \rfloor}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\titleM{}
\begin{abstract}
This document describes the Lambda language used as an intermediate
representation in Moscow ML, ExSML and in Caml Light. It also forms
the basis for Ocaml in a changed form.
\end{abstract}
\listoffixmes
%\tableofcontents
\chapter{Introduction}
This document describes the Lambda-language which is the internal
representation of Moscow ML and ExSML. It is based upon a similar
language in the Ocaml system, although a number of changes has been
made to accomodate SML-specifics.

\chapter{Lambda}

Lambda is an extended lambda-calculus, much like most functional
programming languages use as a basis. By extended, we means that the
language has the lambda-calculus as its base but has been augmented
with additional constructs making certain operations more viable. As
an example, you can define Peano-arithmetic in the pure lambda
calculus -- but this construction is hardly efficient. In Lambda,
integers are provided as a primitive, yielding fast arithmetic. Also,
it is easier to map fast arithmetic down to the underlying machine.

To circumvent problems with alpha-conversion in the traditional
Lambda-calculus, the Lambda language uses De-Bruijn indexing all
over. With De-Bruijn indexing, machine manipulation is easy although
it is harder to read for a human being.

Lambda also contains while-loops and sequencing effectively making
the language into an ``algorithmic functional language''. Contrast with
a pure functional language in which there are no imperative constructs
at all.

First, we introduce some data types used in the Lambda language:

\section{Special Constants}

To facilitate the inclusion of integers, reals, strings, words and
characters in the Lambda language, we use the concept of a
\emph{Special Constant} (SCon). An SCon is a constant in the
traditional sense from a datatype, but it is special in the sense it
is used for primitive values. In the backend a Special constant
is encoded as a \texttt{value}.

Note that the Lambda definition uses an \texttt{StructConstant} as the
constant container rather than an SCon. This is also the reason SCons
are often called Atoms. We will cover StructConstants
later.\fixme{Cover StructConstants Later :)}

\section{BlockTags}

A BlockTag is really a pair of a tag number and a span, embedded in a
\texttt{CONtag} constructor. The tag number is used as a discriminator
so different blocktags can be distinguished from each other. The span
records how many different tag numbers there are in the datatype. For
example, the values ``false'' and ``true'' are encoded as $(0,2)$ and
$(1,2)$ respectively. They are members of a special datatype, bool.

This information is utilized by the pattern match compiler. It knows
when it has outruled the possibility of a match on $span-1$ then it
\emph{must} be the final value. Note that the tag number is not unique
over all different datatypes. We know, due to static type checking
that different tag numbers can not possibly intermingle.

\section{Syntax}

The syntax for the lambda language is as follows:

\newcommand{\ssplit}{\;\vert\;}
\reservestyle{\command}{\mathbf}
\command{let[\;let\;],in[\;in\;],end[\;end\;],case[\;case\;],of[\;of\;],letrec[\;letrec\;],
switch[\;switch],handle[\;handle\;],shandle[\;shandle\;],sfail[\;sfail\;],if[\;if\;],
then[\;then\;],else[\;else\;],shared[\;shared],while[\;while\;],andalso[\;andalso\;],
orelse[\;orelse\;],do[\;do\;],assign[\;assign\;]}
\begin{align*}
  e ::= \quad & \; n \ssplit sc \ssplit e \; (e_1 \; \dotsc \; e_k)
  \ssplit \lambda{}.e
  \ssplit \<let> (e_1 \; \dotsc \; e_k) \<in> e \<end> \ssplit \\
  & \<letrec> (e_1 \; \dotsc \; e_k) \<in> e \<end> \ssplit
  \{p\}(e_1, \dotsc, e_k) \ssplit\\
  & \<case> e \<of> sc_1 \Rightarrow e_1, \; sc_2 \Rightarrow e_2, \;
  \dotsc \ssplit\\
  & \<switch>(a)\; e \<of> bt_1 \Rightarrow e_1, \; bt_2
  \Rightarrow e_2, \; \dotsc \ssplit \\
  & \<sfail> \ssplit e_1 \<shandle> e_2 \ssplit e_1 \<handle> e_2
  \ssplit \\
  & \<if> e_{tst} \<then> e_{true} \<else> e_{false} \ssplit e_1; \; e_2
  \ssplit \\
  & \; \<while> e_t \<do> e \ssplit e \<andalso> e \ssplit e \<orelse> e
  \ssplit unspec \\
  & \<shared>(k) \; e^{*} \ssplit \<assign> i \leftarrow e
\end{align*}
\begin{description}
\item[$n$] A number $n$ refers to the $n$'th enclosing lambda in the
  scope. This is the De-Bruijns indexing on the lambda calculus. That
  is, the $n$ is a natural number referring to a lambda-binder.
\item[$sc$] The $sc$ refers to any special constant. Special
  constants are integers, strings, reals, and so on. They are
  manipulated by using primitive calls. As an example, there is a
  primitive call of arity two which can add integers. Similarly there
  are primitives for string concatenation and so on. In fact all
  operations on structconstants are done by primitives.
\item[$e \; (e_1 \; \dotsc \; e_k) $] Function application in strict
  order. First, $e_1$ through $e_k$ are executed (in that order) and
  when they are reduced to canonical forms they are applied to $e$ in
  order. It is expected that $e$ is a series of lambdas so we can
  apply the arguments.
\item[$\lambda{}.e$] Define a function at this point. In De-Bruijn
  indexing, we do not bind a variable here. Rather we refer to it by
  number.
\item[$\<let> (e_1 \; \dotsc \; e_k) \<in> e \<end>$]
  Let-binding. In some sense, this is redundant as $\<let> x = e_1 \<in>
  e_2 \<end>$ can be encoded as $(\lambda x . e_2) e_1$ (in regular
  Lambda-notation). In Lambda, we have a list of bindings, and we
  refer to them by De-Bruijn indexing as if they had been encoded this
  way.\fixme{Check that this is true}
\item[$\<letrec> (e_1 \; \dotsc \; e_k) \<in> e \<end>$] This is a
  variant of the let-binding in which we have recursion on the
  bindings. That is, a $\<letrec>$ defined lambda abstraction is a
  \emph{fixpoint} in the usual sense if it refers to itself. We make
  the distinction to be able to define recursive fixpoints on
  functions.
\item[$\{p\}(e_1, \dotsc, e_k)$] This construction defines a primitive
  $p$. Primitives does all the hard work on low-level operations:
  integers, strings, reals and so on. Addition, string concatenation
  and like functions are packed as a primitive. Primitives can have
  different arities depending on the primitive being called.
\item[$\<case> e \<of> sc_1 \Rightarrow e_1, \; sc_2 \Rightarrow
    e_2, \; \dotsc$] This operation is the traditional case
  operation. It evaluates the expression $e$ and looks up the
  appropriate matching special constant. Then the corresponding
  expression is evaluated. Example: If $e = sc_3$ then $e_3$ is
  evaluated.
\item[$\<switch>(a)\; e \<of> bt_1 \Rightarrow e_1, \; bt_2
  \Rightarrow e_2, \; \dotsc$] The switch-operation works like the
  case operation, but it acts on constructor tags rather than on the
  value contents. The distinction is made so it can be compiled with
  different strategies.\fixme{Describe the $(a)$ part} The expression
  $e$ is reduced and its tag is looked up. Then the switch is made on
  the contents of this tag.
\item[$\<sfail>, e_1 \<shandle> e_2$] The static exception handling
  is used when exceptions do not escape the current function. Such
  exceptions can be compiled by a goto-statement. Thus one can use
  exceptions locally to mangle the control flow without the costly
  dynamic exceptions.
\item[$e_1 \<handle> e_2$] Dynamic exception handling. Note that
  raising an exception is a primitive function. The $e_2$ expression
  is, in SML, a pattern match construction. This construction is
  pattern-match-compiled down to a switch/case tree and stuffed into
  the place of $e_2$.
\item[$\<if> e_{tst} \<then> e_{true} \<else> e_{false}$] Mere
  If-then-else. It executes either $e_{true}$ or $e_{false}$ depending
  on the outcome of $e_{tst}$. Note that it is expected that $e_{tst}$
  evaluates to a boolean type as we test on the constructor tag of
  false and true to compute which branch to take.
\item[$e_1; \; e_2$] Sequencing. Execute $e_1$ for side-effect only
  before executing $e_2$.
\item[$\<while> e_t \<do> e$] While loop. If $e_t$ evaluates to a true
  value (in the same sense as $e_{tst}$ in the if-then-else), $e$ is
  executed. The whole expression is executed for side-effect only.
\item[$e \<andalso> e, e \<orelse> e$] Short-circuiting logical
  conjunction and disjunction.
\item[$unspec$] An unspecified value. It is used in the construction
  \texttt{val \_ = e} for instance.
\item[$\<shared>(k) \; e^{*}$] This shared rule defines a shared
  lambda construction which is used by the pattern match
  compiler. Rather than building a decision tree, it constructs a
  decision dag. And this shared construct is used for that.
\item[$\<assign> i \leftarrow e$] An assignment construction which
  is used in the backend. How is unknown.
\end{description}
\end{document}
